---
title: "Task 4 Trencin"
author: "Viktor Molodov"
date: "2025/12/02"
format:
  html:
    embed-resources: true
editor: source
---

## Task 4. Building a "Toolkit" for Your Universe (Functions)

In Task 3, you performed a "one-off" analysis of your new dataset. In the real world, you'll want to run that analysis again and again.

The best way to do this is by capturing your code inside **functions**. A function is a reusable "tool" that you build to perform a specific job.

Your mission is to build a "toolkit" of functions that can analyse the dataset you found in Task 3.

**Important:** You must use the **same dataset** and theme you used in Task 3.

### 4.1. The "Greeting" Function (A Simple Start)

Let's start with the basics. Write a simple function that prints a welcome message for your "universe".

1. **Write the function:** It should take at least one argument (e.g., `your_name` or `universe_name`).

2. **Inside the function:** Use `print()` or `cat()` to display a message.

3. **Call the function:** After you define the function, you must call it to see the result.

```r
welcome_to_trencin <- function(your_name) {
  cat("--------------------------------------------------\n")
  cat("      Welcome to the Trenčín Districts Toolkit!\n")
  cat("      Report prepared by:", your_name, "\n")
  cat("--------------------------------------------------\n\n")
}

# Call the function
welcome_to_trencin("Viktor Molodov")
```

### 4.2. The "Loading" Function (A Good Practice)

In Task 3, you wrote code to load and inspect your data. Let's turn that into a function. This is a very common and useful practice.

1. **Write the function:** Name it `load_and_inspect`. It should take one argument: `file_path`.

2. **Inside the function:**

  - Load the `readr` and `dplyr` libraries.

  - Use `read_csv(file_path)` to load the data.

  - Print a message (e.g., "Data loaded successfully.").

  - Print the `glimpse()` of the data.

  - Print the `head()` of the data.

  - **Return** the data frame (use the `return()` command).

3. **Call the function:**

```{r}
library(dplyr)
library(readr)

load_and_inspect <- function(file_path) {

data <- read_csv(file_path, show_col_types = FALSE)

cat("--------------------------------------------------\n")
cat(" Data successfully loaded from:", file_path, "\n")
cat("--------------------------------------------------\n\n")

cat("Dataset contains", nrow(data), "rows and", ncol(data), "columns.\n\n")

cat("Glimpse of the dataset:\n")
print(glimpse(data))

cat("\nFirst 6 rows:\n")
print(head(data))

cat("\n")  # ensures HTML spacing

return(data)
}

# Load Trenčín data

trencin_data <- load_and_inspect("trencin_data.csv")
```

### 4.3. The "Query" Function (The Core of the Task)

This is the most important step. You will turn one of your `dplyr` queries from **Task 3** into a reusable function.

Look back at your Task 3 code. You probably wrote something like this:

  - `my_data |> filter(population > 1000000)`

  - `my_data |> filter(planet_type == "Gas Giant")`

That 1000000 or "Gas Giant" is a "magic number" or "magic string". A function lets you turn it into a parameter.

1. **Choose a query:** Pick one of your `filter()` or `summarise()` queries from Task 3.

2. **Write the function:**
  
  - Give it a descriptive name (e.g., `find_pollution_by_disctrict`, `filter_cities_by_population`).

  - It must take at least two arguments: **data** (for the data frame) and a **parameter** for the value you want to change (e.g., `pop_threshold` or `type_name`).

  - The function should perform the `dplyr` query and `return` the result.

**Example:** Let's say in Task 3 you found population density for each city.

```{r}
population_density_in_district <- function(data, district_name) {

  result <- data %>%
    filter(district == district_name) %>%
    mutate(
      density = population / district_sq_km,
      percent_of_district_population = population / sum(population) * 100,
      deviation_from_mean_density = density - mean(density)
    ) %>%
    arrange(desc(density)) %>%
    select(street, district, population, district_sq_km,
           density, percent_of_district_population,
           deviation_from_mean_density)

  return(result)
}

# Print the table in HTML
print( population_density_in_district(trencin_data, "Zlatovce") )
```

### 4.4. The "Reporting" Function

Let's combine everything. Write a single function that calculates several key statistics for a specific **group** or **category** in your data.

1. **Write the function:** Name it `generate_report`. It should take `data` and a `category_name` as arguments (e.g., a specific city, planet type, or region).

2. **Inside the function**:

    - `filter()` your data for only the `category_name` provided.

    - Calculate 3-4 interesting summary statistics (e.g., `mean()`, `max()`, `min()`, `n()`).

    - Use `cat()` to print a nicely formatted "report" to the console.

**Example:**

```{r}
generate_report <- function(data, district_name) {

dist_data <- data %>% filter(district == district_name)

# Basic counts

street_count <- nrow(dist_data)
total_population <- sum(dist_data$population)

# Population stats

avg_population <- mean(dist_data$population)
min_population <- min(dist_data$population)
max_population <- max(dist_data$population)

# Density stats

densities <- dist_data$population / dist_data$district_sq_km
avg_density <- mean(densities)
max_density <- max(densities)
min_density <- min(densities)
sd_density  <- sd(densities)

cat("\n----------------------------------------------\n")
cat("        DISTRICT REPORT:", district_name, "\n")
cat("----------------------------------------------\n")
cat("Number of streets:", street_count, "\n")
cat("Total population:", total_population, "\n")
cat("Average population:", round(avg_population, 2), "\n")
cat("Min population:", min_population, "\n")
cat("Max population:", max_population, "\n")
cat("\n--- Density Statistics ---\n")
cat("Average density:", round(avg_density, 2), "\n")
cat("Min density:", round(min_density, 2), "\n")
cat("Max density:", round(max_density, 2), "\n")
cat("Standard deviation of density:", round(sd_density, 2), "\n")
cat("----------------------------------------------\n\n")
}

# Generate two district reports

print( generate_report(trencin_data, "Zlatovce") )
print( generate_report(trencin_data, "Centrum") )
```